RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-08-13 background.tex
fix conflicts

These are the types of the questions that we cannot answer by just looking at GRL models. The model in Figure~\ref{fig:example-small} does not contain information about discussions that led to the model, such as clarification steps for the naming, or alternatives that have been considered for the relationships. The idea behind the original GRL specification is that beliefs can be used to capture such design rationales that make later justification and review of a model easier. However, beliefs cannot be connected to links - this makes answering the third and fourth question above impossible. Furthermore, beliefs are after-the-fact design rationales and do not capture the types of questions given above. Finally, the jUCMNav tool does not consider beliefs in the evaluation algorithms, and thus there is no formal connection between goal models and their underlying beliefs and opinions. For a more detailed comparison of our framework with GRL beliefs, see Section~\ref{sect:discussions:relatedwork}.

>>>>

These are the types of the questions that we cannot answer by just looking at GRL models. The model in Figure~\ref{fig:example-small} does not contain information about discussions that led to the model, such as clarification steps for the naming, or alternatives that have been considered for the relationships. The idea behind the original GRL specification is that so-called \emph{belief} elements can be used to capture such design rationales that make later justification and review of a model easier. In the original GRL specification, however, the exact semantics of these belief elements and the ways in which they should be used remain unclear. Furthermore, belief elements cannot be attached to links between elements, which makes answering the fourth and fifth question above impossible. We further discuss belief elements in light of our proposed argumentation-based framework in Section~\ref{sect:discussions:relatedwork}.

### 2018-08-13 discussion.tex
fix conflicts

\paragraph{Goal Modeling} Argumentation has been included -- both explicitly and implicitly -- in existing goal modeling languages. For example, the belief element in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} is meant to capture the rationales behind the inclusion of goals and tasks in the model. Furthermore, relations between elements in a goal model also provide justifications: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative subgoal or task in a goal model is then an explicit argumentative move in the discussion. In this sense, a goal model provides a justification for itself, particularly if we include belief elements for extra design rationalization. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models, it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are acceptable given potentially contradictory opinions of stakeholders.

There are several contributions to the literature that relate argumentation-based techniques with goal modeling. The first line of work is by Bagheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract argumentation frameworks (see Definition~\ref{def:argumentation-framework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics \cite{Dung1995}. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled as such, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

@@

\paragraph{Beliefs in GRL} As mentioned already in the introduction, it is possible in GRL to rationalize models
through so-called \emph{belief} elements. These beliefs are elements that can be attached to an IE with a \emph{belief link}. The content of a belief is a textual description, which can explain, for instance, the rationale for including an IE in the resulting goal model. There are several important differences between GRL beliefs and RationalGRL arguments.

First of all, beliefs in GRL are not used in any of the algorithms to compute the satisfiability of the IEs. This means that a GRL belief can be understood as metadata that can be attached to an element, but it does not have semantics. Argument in RationalGRL, in contrast, play an important role in forming the resulting goal model. If an IE is defeated by an argument, it is not part of the resulting goal model.

Secondly, arguments in RationalGRL can not only provide rationale for IEs in the resulting goal model, but they can also provide rational for those that are rejected, and they can be attached to links as well. This means that Rational arguments are strictly more expressive than GRL beliefs.

Finally, the formation of arguments and counterarguments in RationalGRL is guided by critical questions, which stimulate discussions between stakeholders. This is not possible for GRL beliefs.

>>>>

\paragraph{Goal Modeling} Argumentation has already been included implicitly in existing goal modeling languages. At its core, goal modeling is a process of argumentation: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative subgoal or task in a goal model is then an explicit argumentative move in the discussion. In this sense, a goal model provides a justification for itself. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models, it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are \emph{acceptable} (cf. Definition 14} given potentially contradictory opinions of stakeholders.

More explicit rationales for goal models are included in the original GRL specification \cite{} in the form of \emph{belief} elements, which can be attached to other intentional elements (goals, tasks) using the different links (positive contribution, negative contribution) of GRL. The content of a belief is a textual description, which can explain, for instance, the rationale for including a goal in a model. In this regard, they seem like arguments or reasons for the different elements in a goal model. However, in the literature the precise usage and semantics of belief elements in GRL remain unclear: according to the metamodel \cite{•} they can be connected to other intentional elements (including beliefs) through any type of link, but in the jUCMNav tool there is a separate \emph{belief link} for connecting beliefs to goals and tasks only, and the beliefs are not used in any of the algorithms to compute the satisfiability of goals and tasks. We are also not aware of many examples of the usage of belief elements in the literature. So what it seems like is that GRL belief elements can be understood as metadata that can be attached to an element

Ambiguities notwithstanding, there are further advantages of using RationalGRL over GRL with belief elements. For example, in none of the GRL specifications is it possible to attach beliefs to links between other elements so it is not possible, for example, to provide a rationale for why some tasks positively contributes to a goal. Furthermore, the RationalGRL arguments follow clear schemes with associated critical questions explicitly tailored towards reasoning about goals and actions or tasks, while beliefs in GRL offer no further guidance as to what kinds of rationales can be given for elements in a goal model. Furthermore, beliefs in GRL seem not to influence the goal model in any way. Arguments in RationalGRL, in contrast, play an important role in forming the resulting goal model: if an element is defeated by an argument, it is not part of the resulting goal model.

Considering beliefs, an interesting avenue for further research would be to look at to connections between argumentation and belief revision \cite{alchourron1985logic}, and how these can play a role in RationalGRL and requirements engineering in general. While there is formal work in Artificial Intelligence on the connections between argumentation and belief revision (see e.g. \cite{falappa2009belief}, there is very little work on applying the combination to requirements engineering. The only work we are aware of that deals with his subject is by Gervasi and Zowghi \cite{gervasi2005reasoning}, who use a logical framework for belief revision to detect inconsistencies in software specifications and then revise those specifications so that the inconsistencies are removed. They do not explicitly connect their framework to goal models of goal modelling, however, and do not connect their framework to specific argumentation using schemes in the way RationalGRL does.

\paragraph{Goal Modeling and Argumentation}

There are several contributions to the literature that explicitly relate argumentation-based techniques with goal modeling. The first line of work is by Bagheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract argumentation frameworks (see Definition~\ref{def:argumentation-framework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics \cite{Dung1995}. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled as such, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

@@
